---
title: "Class 9 Brest cancer analysis with PCA and clustering"
author: "Barry Grant"
date: "2/7/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read our data

```{r}
# Save your input data file to a new 'data' directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data)
```

Lets see the first 6 rows of data

```{r}
head(wisc.df)
```

How many samples (i.e. patients) are in this data-set?

```{r}
nrow(wisc.df)
```

Next use as.matrix() to convert the other features (i.e. columns) of the data (in columns 3 through 32) to a matrix. Store this in a variable called wisc.data.

```{r}
# Convert the features of the data: wisc.data
wisc.data <- as.matrix( wisc.df[,3:32] )
```

Add rownames to our new matrix of data

```{r}
rownames(wisc.data) <- wisc.df$id 
```

Finally, setup a separate new vector called diagnosis to be 1 if a diagnosis is malignant ("M") and 0 otherwise. Note that R coerces TRUE to 1 and FALSE to 0.

How many cancer (M) and non cancer samples do we have in our data-set?
```{r}
table(wisc.df$diagnosis)
```


```{r}
diagnosis <- as.numeric( wisc.df$diagnosis == "M" )
```


Q1. How many observations are in this dataset?
```{r}
nrow(wisc.data)
```

Q2. How many variables/features in the data are suffixed with _mean?
```{r}

inds <- grep("_mean", colnames(wisc.data))

length(inds)
```


Q3. How many of the observations have a malignant diagnosis?

```{r}
sum(diagnosis)
```


## Lets do PCA

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp( wisc.data, scale=TRUE )
summary(wisc.pr)
```

Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?


## Lets see plots of our results

First a standard biplot

```{r}
biplot(wisc.pr)
```

We need to make our own PCA plot

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2])
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=diagnosis+1)
```


### Variance captured in each PC

This info is in the $sdev component of our PCA result
```{r}
varance <- wisc.pr$sdev^2
# Variance explained by each principal component: pve
pve <- round( (varance/sum(varance)) * 100, 1)
```


```{r}
plot(pve, type="o")
```

```{r}
barplot(pve)
```

Make the plot a bit fancy :-o


```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2) )

```


## Hierarchical clustering of cancer case data

For hierarchical clustering we need a few things
1. Distance matrix **dist()** function
2. The **hclust()** function.
3. Use the **cutree()** function to find cluster membersip vector

```{r}
# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)
```


```{r}
round( apply(wisc.data, 2, sd), 1)

```


```{r}
round( apply(data.scaled, 2, sd), 1)

```

Looks like we need scalling!!

Calculate the (Euclidean) distances between all pairs of observations in the new scaled dataset and assign the result to data.dist.

```{r}
data.dist <- dist(data.scaled)
```

```{r}
wisc.hclust <- hclust(data.dist, method="complete")
```

```{r}
plot(wisc.hclust)
```


### Cluster in PCA space....

For clustering we need?
1. Distance matrix
2. Clustering function
3. Cutree

This was our PCA result of PC1 vs PC2
```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=diagnosis+1)
```

```{r}
pc.dist <- dist( wisc.pr$x[,1:2] )
pc.hclust <- hclust( pc.dist, method="ward.D2")
plot(pc.hclust)
```

```{r}
grps3 <- cutree(pc.hclust, k=3)
table(grps3)
```

```{r}
table(grps3, diagnosis)
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,2], col=grps3)
```

## Prediction

We will use the predict() function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis+1)
points(npc[,1], npc[,2], col="blue", pch=15, cex=3)
```

